### 2022-10-08 manuscript/manuscript.tex
more theory...

\subsection*{Overview of Theory}

Here, I step through a quick derivation of \textcite{Santiago1998-bs}.
Throughout, we assume random mating, hermaphroditic individuals, and a constant
population size.

The change in a neutral allele's frequency in one generation can be partitioned
into the three sources of stochasticity: the random associations with fitness
backgrounds (i.e. \emph{draft}), the non-heritable randomness in family size,
and the Mendelian noise from heterozygotes segregating. If we let $x_{0,i} \in
\{0, \nicefrac{1}{2}, 1\}$ be the frequency of neutral alleles individual $i$
in generation 0 carries, we can partition the random neutral allele frequency
of the population ($x_t$ without the individual index) into the underlying
stochastic causes,

@@

\begin{align}
  \var(\Delta p_1) &= K_1 + H_1 \\
  &\approx \frac{p_0(1-p_0)}{2N} \frac{\var(k)}{4} + \frac{p_0(1-p_0)}{2N} \frac{1}{2} \\
                   &\approx \frac{p_0(1-p_0)}{2N}\left(\frac{\var(k)}{4} + \frac{1}{2}\right)
\end{align}

@@

\begin{align}
  \V(x_2) &= \E\left[(x_2 - x_0)^2\right] \\
          &= \E_1\left[\E_2\left[((x_2-x_1) + (x_1 - x_0))^2 | x_1\right]\right] \\
          &= \E_1\left[\E_2\left[(\Delta x_2 + \Delta x_1)^2 | x_1\right]\right] \\
          &= \E_1\left[\E_2\left[{\Delta x_2}^2|x_1\right]\right] + 2\E_1\left[\E_2[\Delta x_2 \Delta x_1 | x_1]\right] + \E_1\left[ {\Delta x_1}^2\right] \\
          &= \E_1\left[\E_2\left[{\Delta x_2}^2|x_1\right]\right] + \E_1\left[ {\Delta x_1}^2\right] + C_{1,2} \\
          &= \frac{p_0(1-p_0)}{2N}\left(1 - \frac{1}{2N}\right) + \frac{p_0(1-p_0)}{2N} + C_{1,2}
\end{align}

@@

>>>>

\subsection*{Overview of Quantitative Genetic Linked Selection Theory}

Here, we step through a quick derivation of \textcite{Santiago1998-bs}.
Throughout, we assume random mating, hermaphroditic individuals, and a constant
population size. The change in a neutral allele's frequency in one generation
can be partitioned into the three sources of stochasticity: the random
associations with fitness backgrounds (i.e. \emph{draft}), the non-heritable
randomness in family size, and the Mendelian noise from heterozygotes
segregating. If we let $x_{0,i} \in \{0, \nicefrac{1}{2}, 1\}$ be the frequency
of neutral alleles individual $i$ in generation 0 carries, we can partition the
random neutral allele frequency of the population ($x_t$ without the individual
index) into the underlying stochastic causes,

@@

\begin{align}
  \var(\Delta x_1) &= \var(K_1) + \var(H_1) \\
  &\approx \frac{p_0(1-p_0)}{2N} \frac{\var(k)}{4} + \frac{p_0(1-p_0)}{2N} \frac{1}{2} \\
                   &\approx \frac{p_0(1-p_0)}{2N}\left(\frac{\var(k)}{4} + \frac{1}{2}\right)
\end{align}

@@

\begin{align}
  \label{eq:var_wf}
  \V(x_2) &= \E\left[(x_2 - x_0)^2\right] \\
          &= \E_1\left[\E_2\left[((x_2-x_1) + (x_1 - x_0))^2 | x_1\right]\right] \\
          &= \E_1\left[\E_2\left[(\Delta x_2 + \Delta x_1)^2 | x_1\right]\right] \\
          &= \E_1\left[\E_2\left[{\Delta x_2}^2|x_1\right]\right] + 2\E_1\left[\E_2[\Delta x_2 \Delta x_1 | x_1]\right] + \E_1\left[ {\Delta x_1}^2\right] \\
          &= \E_1\left[\E_2\left[{\Delta x_2}^2|x_1\right]\right] + \E_1\left[ {\Delta x_1}^2\right] + C_{1,2} \\
          &= \frac{p_0(1-p_0)}{2N}\left(1 - \frac{1}{2N}\right) + \frac{p_0(1-p_0)}{2N} + C_{1,2}
\end{align}

@@

Now, we turn our attention to the case where there is heritable fitness
variation, which leads to non-zero covariance terms $C_{i,j}$. These terms
emerge when $\var(k)$ has a heritable component of fitness that can be
transmitted along with the neutral allele, thereby affecting the neutral
allele's trajectory in later generations. If we write the offspring number into
heritable and non-heritable components, $k_i = 2f_i + \varepsilon_i$, then
$\var(k_i) = 4 V_h + V_n$. We can further decompose the $K_1$ term into $K_1 =
S_1 + D_1$,

\begin{align}
  \var(\Delta x_1) &= \var(S_1) + \var(D_1) + \var(H_1) \\
  &\approx \frac{p_0(1-p_0)}{2N} V_h + \frac{p_0(1-p_0)}{2N} \frac{V_n}{4} + \frac{p_0(1-p_0)}{2N} \frac{1}{2}. \\
\end{align}

(c.f. \cite{Santiago1995-hx} equation 11).

To see how the covariance terms accumulate, consider $\var(x_3 - p_0)$. Note
that $\E(S_t) = \E(D_t) = \E(H_3) = 0$, since which neutral allele we track is
arbitrary, so by symmetry the expected change is zero. Then,

\begin{align}
  \label{eq:var_x}
  \var(x_3 - p_0) &= \E\left[ \left(S_1 + D_1 + H_1 + S_2 + D_2 + H_2 + S_3 + D_3 + H_3 \right)^2 \right] \\
                  &= \E({S_1}^2) + \E({S_2}^2) + \E({S_3}^2)  \\
                  &+ 2\E(S_1 S_2) + 2\E(S_1 S_3) + 2\E(S_2 S_3) \\
                  &+ \E({D_1}^2) + \E({H_1}^2) + \E({D_2}^2) + \E({H_2}^2) + \E({D_3}^2) + \E({H_3}^2)
\end{align}

The covariance terms $\E(S_i S_j)$ for $j > i$ represent the expected
neutral allele frequency change from a neutral allele becoming associated with
a fitness background in generation $i$ and that fitness association persisting
until generation $j$. Let us consider $S_1$ and $S_2$. We have,

\begin{align}
  S_1 = \frac{1}{2N} \sum_{i=1}^N x_{0,i}(f_{0,i} - 1) = \cov(x_{0,i}, f_{0,i}) \\
  S_2 = \frac{1}{2N} \sum_{i=1}^N x_{1,i}(f_{1,i} - 1) = \cov(x_{1,i}, f_{0,i})
\end{align}

which are chance covariances (across the population, \emph{not} evolutionary
replicates) created by the random sorting of neutral alleles and fitness into
individuals, \emph{each generation}. These covariances are equivalent to the
Robertson-Price equation \parencite{Robertson1966-fs,Price1970-si}, which
predict the change in neutral frequency due to heritable fitness (and likewise
with the change due to non-heritable factors).

Considering the underlying haplotypes that lead to $x_{0,i}$ and $f_{0,i}$ can
give us an equation for the dynamics of the $\E(S_i S_j)$ (for $j > i$) terms
over evolutionary replicates. Let us partition the allele frequency and fitness
by into the average of the paternal contributions. The neutral allele frequency
per gamete is either 0 or 1, and we assume the fitnesses across gametes are
additive. Then, as long as there is random mating, we can simplify the
associations in a diploid by looking at a single gamete,

\begin{align}
  S_1 = \cov(x_{0,i}, f_{0,i}) &= \cov\left(\frac{x_{0,i}' + x_{0,i}''}{2}, f_{0,i}' + f_{0,i}''\right)  \\
                               &= \frac{1}{2}\left(\cov(x_{0,i}', f_{0,i}' + f_{0,i}'') + \cov(x_{0,i}'', f_{0,i}' + f_{0,i}'')\right)  \\
                               &= \cov(x_{0,i}', f_{0,i}' + f_{0,i}'')  \\
                               &= \cov(x_{0,i}', f_{0,i}') + \cov(x_{0,i}', f_{0,i}'')  \\
                               &= S_1' + S_1''
\end{align}

which follows from symmetry, since it does not matter which gamete carrying the
neutral allele we track (c.f. \cite{Santiago1998-bs} p. 2107). The primes now
indicate whether the selected locus is on the same gamete as the neutral
allele (single prime), or the homologous gamete (double prime).

After meiosis, a fraction $1-r$ of these associations between the tracked
neutral neutral allele $x_{0,1}'$ and fitness background $f_{0,i}'$ persist,
and a fraction $r$ disassociate its currently coupled background and create an
association with the homologous fitness background, $f_{0,i}''$. When linkage
is tight, this latter term can be ignored as we will do here, though it does
impact background levels of neutral diversity \parencite{Santiago1995-hx}.
Additionally, the fitness effects of the selected background may change in
later generations, in complex ways \parencite{Barton1986-yh,Turelli1990-kd}.
\textcite{Santiago1995-hx} assume an equilibrium level of fitness variation
$V_h$, with the fitness variance associated with a particular background
decaying at a simple geometric decay at rate $Z$ each generation, which is
sufficient for background selection. An approximate rate of the decay can be
worked out for the particular selective system. In general, this can be a much
more complicated function; \textcite{Buffalo2019-qs} show that even fluctuating
selection can be accommodated.

Now, the associations created in generation 1 and that persist in future
generations as (ignoring $S_i''$ terms) follow the pattern

\begin{align}
  S_{2}' &= S_1 Z(1-r) \\
  S_{3}' &= S_1 Z^2(1-r)^2 \\
  S_{4}' &= S_1 Z^3(1-r)^3.
\end{align}

Then, the cumulative effect of the fitness associations created in generation
$1$ can be written as,

\begin{align}
  S_1 Q_t &= S_1 + S_2' + S_3' + \ldots + S_t' \\
  Q_t &= 1 + \sum_{i=1}^t r^i (1-Z)^i \\
  Q_\infty &= \frac{1}{1-r (1-Z)}
\end{align}

which shows the long reach of heritable fitness factors in altering allele
frequency through the generations. In terms like $\E(S_1 S_2)$, a fraction
$S_1'$ of the heritable allele frequency change $S_2$ were \emph{previously}
built associations between the neutral allele and a fitness background from
$S_1$. As long as the direction of selection is the same, the expected change
of this fraction of associations formed in the first generation to later
generations 2 and 3 would be,

\begin{align}
  \E(S_1 S_2) &= \E(S_1^2) Z(1-r) = \E(S_1^2) Q_2^2 \\
  \E(S_1 S_3) &= \E(S_1^2) Z^2(1-r)^2 = \E(S_1^2) Q_3^2.
\end{align}

With these covariance terms, we now have a full model for $\var(p_t - p_0)$. We
will look at the case for $t=2$. As shown in equation XXX, the second moment of
every term is a function of the variance in neutral allele frequency in that
generation, $x_t(1-x_t)$. This variance does not stay its initial value of
$x_0(1-x_0)$; it decays each generation due to coalesce from the drift and
selection processes modeled by this approach. The rate at which this decay
happens is the effective population size each generation implied by this model. 
We write (this differs from the original \cite{Santiago1995-hx} derivation),

\begin{align}
  \var(x_2 - p_0) &= \E({S_1}^2) + 2\E(S_1 S_2) + \E({S_2}^2) \\
                  & + \E({D_1}^2) + \E({H_1}^2) + \E({D_2}^2) + \E({H_2}^2) \\
  \frac{\var(x_2 - p_0)}{p_0(1-p_0)} &= \frac{V_h}{2N_{e,1}} + \frac{V_h Q_2^2}{2N_{e,2}}\left(1-\frac{1}{2N_{e,1}}\right)  +  \frac{V_h}{2N_{e,2}}  \left(1-\frac{1}{2N_{e,1}}\right)  \\ 
                                     &+ \frac{V_n}{8N_{e,1}} + \frac{1}{4N_{e,1}} + \frac{V_n}{8N_{e,2}}  \left(1 - \frac{1}{2N_{e,1}}\right) 
+ \frac{1}{4N_{e,2}} \left(1 - \frac{1}{2N_{e,1}}\right).
\end{align}

A general equation of this is,

\begin{align}
  \frac{\var(x_t - p_0)}{p_0(1-p_0)} &= \sum_{i=1}^\infty \left(\frac{V_h Q_i^2}{2 N_{e,i}} + \frac{V_n}{8N_{e,i}} + \frac{1}{4N_{e,i}} \right)\prod_{t=1}^{i-1} \left(1-\frac{1}{2 N_{e,t}}\right) \\
\end{align}

In the original derivation of \textcite{Santiago1995-hx}, they set each
$N_{e,t}$ term to a single $N_e$ for which they solve for.

This is an asymptotic $N_e$, but this does not accurately describe the levels
of diversity (\cite{Santiago1998-bs} equation 16, \cite{Santiago2016-mu}
equation 4).

\begin{align}
  \V(x_t) &= \V(x_1) \left(1 + \sum_{i=1}^t \left(1 - \frac{1}{2N}\right)^{i} \right) + \sum_{i \ne j} C_{i,j}
\end{align}
